---
title: "week07_exercises.qmd"
output: html_document
---

# [Yeast cell cycle / transcriptomics exercise #2]{.underline}

# PART 1: Network analysis of the Alpha Factor Arrest data

## Network analysis

The final part of the analysis of the alpha-factor arrest data set we started in the previous exercise, is to map it onto the Yeast protein-protein interaction network we worked with in week 4.

## Load Libraries

```{r}
library(tidygraph)
library(igraph)
library(ggraph)
library(ggplot2)
```

**TASK: reload base session, prepare Excel data for import**

```{r}
load("/home/projects/22140/exercise4.Rdata")
load("/home/projects/22140/exercise6.Rdata")
# Calculate log2fc from expression
expr$log2fc <- log2(expr$GSM287992/expr$GSM287991)
# Add log2fc to your node attribute table
node_attributes_updated <- merge(x = node_attributes, y = expr[!duplicated(expr$SysName),c(1,6)], by.x = "ID", by.y = "SysName", all.x = TRUE)
# If you haven't worked with the "merge" command, take a moment to understand the line above. It's a super useful command
```

**TASK: Visualize the Log2(FC)**

-   Plot the network using ggraph with the following mappings:

-   Color nodes by log2FC

-   Use the "scale_color_gradient2" argument to ggraph to color down regulated nodes blue, (mid color gray), and up regulated red

-   Set node shape based on whether the log2fc is lower than -2 or higher than 2. Hint:

```{r}
g <- graph_from_data_frame(d = interactions, 
                           directed = FALSE, 
                           vertices = node_attributes_updated)

ggraph(g) +
  geom_edge_link() +
  geom_node_point(aes(color = log2fc, shape = abs(log2fc)>2), size = 3) +
  scale_color_gradient2(low = "red", mid = "gray", midpoint = 0, high = "blue")

```

The KAR genes are involved in the karyogamy process (also discussed in the answers for last weeks exercise), and it makes good sense that they are overexpressed in the alpha-factor arrested genes. (Alpha-factor triggers the mating response, which in turn prepares for a fusion of the nuclei of the A- and alpha-cells – this is the process known as karyogamy).

```{r}
expr[grepl(pattern = "kar", x = expr$PopName, ignore.case = TRUE),]

```

**Clusters of Regulation**: You can visually identify groups of genes (nodes) that are connected and share similar regulation trends (indicated by color and shape).

-   **Color**: The gradient from blue to red indicates the direction and magnitude of the log2 fold change (log2fc), with blue being highly upregulated and red being highly downregulated.

-   **Shape**: Triangles represent genes with significant regulation (`abs(log2fc) > 2`), while circles indicate less significant regulation.

The next task is to have a look at the network and try to interpret the results. Here it should be noted that the cells have been arrested in what may be a bit "boring" part of the cell cycle (late G1), but we can still make a few interesting observations.

**REPORT QUESTION #1: Inspect the network**

-   Can you find any clusters (even small ones) with several genes being regulated in the same direction?

-   Discuss the biological meaning of this within the group (note that not all clusters are super-easy to interpret).

-   Which of the KAR genes are regulated?

    -   KAR5, KAR3 and KAR4

```{r}
expr[grepl(pattern = "kar", x = expr$PopName, ignore.case = TRUE),]

```

# PART 2: Arrest and Release time-series experiment

## The "alpha-30/alpha-38" arrest and release experiments

For the first part of the exercise we'll be working an alpha-factor arrest-and-release experiment (the "**alpha-30/alpha38**" experiment from the Breeden lab). Briefly, the experimental set-up is as follows:

-   **ARREST:** The culture was arrested using alpha-factor (as we have seen before).

-   **RELEASE:** When most of the cells had been arrested in cell cycle, the cells were spun down, and re-suspended in fresh media (thus removing the alpha-factor).

-   **SAMPLING:** small samples were collected from the culture at 5-minute intervals following release (and experimental tricks were used to quickly kill the cells and protect the RNA).

-   **ARRAY ANALYSIS:** for each timepoint the synchronized cells were compared to an asynchronous culture on a **two color array** (competitive hybridization) using Cy3 and Cy5 labeling.

    -   **DYE SWAP:** The experiment was carried out twice, once with CASE (Cy3) vs. CONTROL (Cy5) and once with CASE (Cy5) vs. CONTROL (Cy3) - this is done to eliminate technical biases in the labeling process.

### **Understanding the data**

**TASK: find the "alpha-30" experiment in GEO"**

-   Search for the accession ID: GSE4987

**REPORT QUESTION #2:**

-   Find the title of the publication describing the experiment.

    -   "The Forkhead transcription factor Hcm1 regulates chromosome segregation genes and fills the S-phase gap in the transcriptional circuitry of the cell cycle"

-   How many arrays (measurements) are associated with the experiment?

    -   50 (samples)

**CLEANED UP DATA:**

From the data available for download at GEO (the MATRIX file mentioned above), we have prepared an extract (and very slight reformatting) of the data we need for this exercise.

```{r}
load("/home/projects/22140/exercise7.Rdata")
```

-   The data frame "alpha30_38" contains the data in long format. Take a minute to explore the data frame. The log2fc is the fold change between the two conditions case and control. In the alpha30 experiment, fold change is case = Cy3 / control = Cy5. In alpha38 the dyes were swapped and hence control = Cy3 / case = Cy5

### **Estimating the inter-division time**

Before we move on with the analysis of biology described by the data, we need to have a better understanding of how the **time series** relates to the cell cycle. We'll start out by estimating the **inter-division time** (number of minutes it takes for a full cell cycle).

\
This can be done by simply plotting the **expression profiles** of a few genes that we expect to follow a **cyclic pattern** in the data set, for example:

```         
 
RAD53 [YPL153C] G1 (mid)  DNA repair/cell cycle arrest
CLN1  [YMR199W] G1/S1     G1-cyclin - controls entry to the S-phase
HTB2  [YBL002W] S (mid)   Histone H2B - histones are needed for the new chromosomes
CLB1  [YGR108W] G2 (mid)  B-type cyclin
ASH1  [YKL185W] M (late)  Transcriptional regulator (during anaphase)
```

**TASK:** below is a hint how to use ggplot to plot x = timepoint and y = log2fc

```{r}
# all in one
df <- alpha30_38[alpha30_38$gene %in% c("YPL153C", "YMR199W", "YBL002W", "YGR108W", "YKL185W") & alpha30_38$experiment == "alpha30",]
ggplot(df, aes(x = timepoint, y = log2fc, color = gene)) +
  geom_point() +
  geom_line() +
  ggtitle("alpha30")

# or one at the time if you prefer
ggplot(df, aes(x = timepoint, y = log2fc, group = gene)) +
  geom_point() +
  geom_line() +
  facet_wrap(~gene) +
  ggtitle("alpha30")
```
